---
title: "Week 06 Assignment"
author: "Gayathri Pavani Guntamukkala"
date: "2025-03-02"
output:
  html_document:
    df_print: paged
  pdf_document: default
---
    


* We must choose eight Federally Qualified Health Centers (FQHCs) in Missouri to establish specialized mental health services in this assignment. The objective is to establish the best possible health service accessibility and geographic coverage for Missouri residents. The achievement of this goal requires the use of a genetic algorithm (GA).

First lets install required libraries.
```{r}
#if (!requireNamespace("GA", quietly = TRUE)) {
 # install.packages("GA")
#}

```

* It checks if the GA package (used for genetic algorithms) is installed. If not, it installs the package. The requireNamespace("GA", quietly = TRUE) function verifies the package's presence without loading it, and install.packages("GA") installs it if missing.

```{r}
required_packages <- c("sp", "raster", "sf", "terra", "pkgbuild", "devtools")

# Check for and install any missing packages
for (package in required_packages) {
   if (!require(package, character.only = TRUE)) {
     install.packages(package)
   }
 }
```
Installing supporting libraries for rgdal

```{r}
url <- "https://download.r-forge.r-project.org/bin/windows/contrib/4.4/rgdal_1.6-7.zip"
install.packages(url, type="source", repos=NULL)
```

* As The rgdal package is not available on CRAN for my specific R version, I am directly downloading and installing the ZIP file of rgdal from a non-CRAN source.

```{r}
library(data.table)
library(geosphere)
library(rgdal)
library(GA)
library(purrr)
```
Loading required libraries.

```{r}

## Load the population data
pop.data <- data.table::fread("Mo_pop_Sim.csv")
str(pop.data)

small.data <- pop.data[sample(1:nrow(pop.data),
                       size = round(nrow(pop.data) * 0.0001),
                       replace = F)]

## Load the FQHC data
data_path <- 'MO_2018_Federally_Qualified_Health_Center_Locations/MO_2018_Federally_Qualified_Health_Center_Locations/'

fqhc.data <- readOGR(dsn = data_path, layer = 'MO_2018_Federally_Qualified_Health_Center_Locations')

## Convert the data to a data.table for easier manipulation
fqhc.data <- data.table(as.data.frame(fqhc.data))

str(fqhc.data)
```

* Now we are loading population data from Mo_pop_Sim.csv using fread() and extracts a small random subset (0.01%). And then we are loading Federally Qualified Health Center (FQHC) data from a specified path using readOGR(), converts it into a data.table.

*Now its time to select one approach and continue with it. I choose approach 2 and the I selected this metric because it directly identifies the maximum resident access to services which constitutes a central element of healthcare accessibility. We choose locations with the highest resident numbers within a 30-mile radius to guarantee maximum access for people.

```{r}
fqc <- fqhc.data[, .(Longitude, Latitude)]

poc <- small.data[, .(long, lat)]

```

* Lets continue with the code. Now we extract longitude and latitude coordinates for both FQHC and population datasets.

```{r}
# Calculate population within 30-mile radius for each FQHC
radius_miles <- 30

radius_meters <- radius_miles * 1609.34  # Convert miles to meters

pop_counts <- numeric(nrow(fqc))  # Initialize vector to store population counts

for (i in 1:nrow(fqc)) {
  fc <- fqc[i, ]
  distances <- distHaversine(matrix(c(fc$Longitude, fc$Latitude), ncol=2),
                             matrix(c(poc$long, poc$lat), ncol=2))
  pop_counts[i] <- sum(distances <= radius_meters)
}

fqhc.data$pop_in_30mi <- pop_counts
fqhc.data$estimated_pop_in_30mi <- fqhc.data$pop_in_30mi * (1/ 0.0001)
```

* The above cell code is really important as we are trying to  calculate the population within a 30-mile radius for each Federally Qualified Health Center (FQHC). 

* It iterates through each FQHC, computes distances to all population points using the Haversine formula, and counts those within the specified radius (30 mil). The results are stored in fqhc.data, with an estimated population adjustment using a scaling factor (0.0001). The scaling factor adjusts for sampling since "small.data" represents only a fraction of the full population. Multiplying by 1/0.0001 estimates the actual population within a 30-mile radius.

```{r}
eres <- function(fc) sum(distHaversine(as.matrix(fc), as.matrix(poc)) <= radius_meters)

# Genetic Algorithm Fitness Function
fitness_function <- function(indices) sum(sapply(indices, function(i) eres(fqc[i, ])))

# Run Genetic Algorithm
ga_result <- ga(
  type = "permutation", fitness = fitness_function,
  lower = 1, upper = nrow(fqc), popSize = 50,
  maxiter = 10, run = 10, pmutation = 0.2,
  crossover = gaperm_oxCrossover, elitism = 2
)

# Extract best solution
best_fqhc <- fqhc.data[ga_result@solution[1, ], .(OBJECTID, Facility, City, estimated_pop_in_30mi)]
print(best_fqhc)

```

First we started with doing same as previous cell which is computing distances to all population points and made it into a simple function now. Next is the fitness_function() which evaluates a set of FQHC locations based on their population coverage. And genetic algorithm (GA) optimizes the selection of FQHCs to maximize coverage, using permutation-based evolution with crossover and mutation strategies. The best solution is extracted and displayed.

Given results show that these top 8 locations are the optimal choices for introducing specialized mental health services.

As explained before the rationale for this choice is that it directly measures the potential number of residents who can access the services. This analysis or method assumes that the FQHC locations in the dataset are accurate and up-to-date. It also assumes that a 30-mile radius is a reasonable or easy measure of accessibility, as residents are likely to seek services within this distance.

The goal is to maximize population coverage, ensuring efficient allocation of mental health funds and improved access while enhancing service sustainability. So this method will ensure  efficient service.


**THANK YOU!!**

